Supplementary materials for:\\ A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}

### 2021-01-14 article_format/extensions_format_revision.tex
Started major changes for resubmission

@@

\CommentTok{# Run the FE model in plm }
\NormalTok{fe1 <-}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x, }
           \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"within"}\NormalTok{, }
           \DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(fe1)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Generate the cluster means for x per id }
\NormalTok{clusterMeanx <-}\StringTok{ }\KeywordTok{aggregate}\NormalTok{(df}\OperatorTok{$}\NormalTok{x, }\DataTypeTok{by =} \KeywordTok{list}\NormalTok{(df}\OperatorTok{$}\NormalTok{id), }\DataTypeTok{FUN =}\NormalTok{ mean)}
\CommentTok{# Rename the columns }
\KeywordTok{names}\NormalTok{(clusterMeanx) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"xbar"}\NormalTok{)}

\CommentTok{# Add the cluster means back into df  }
\NormalTok{df <-}\StringTok{ }\KeywordTok{merge}\NormalTok{(df, clusterMeanx, }\DataTypeTok{by =} \StringTok{"id"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe2 <-}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar, }
           \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"random"}\NormalTok{,}
           \DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(fe2)}
\end{Highlighting}
\end{Shaded}

@@

\CommentTok{# Run the mixed model in lmer with the cluster means for x }
\NormalTok{mixed1 <-}\StringTok{ }\KeywordTok{lmer}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{id), }\DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(mixed1)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Print the variance components for the plm model  }
\KeywordTok{print}\NormalTok{(}\KeywordTok{ercomp}\NormalTok{(fe2), }
      \DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Print the variance components for the lmer model }
\KeywordTok{print}\NormalTok{(}\KeywordTok{VarCorr}\NormalTok{(mixed1), }
      \DataTypeTok{comp =} \KeywordTok{c}\NormalTok{(}\StringTok{"Variance"}\NormalTok{, }\StringTok{"Std.Dev"}\NormalTok{), }
      \DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem_fullyrelaxed <-}\StringTok{ '}
\StringTok{# Define individual effects variable }
\StringTok{a =~ y1 + y2 + y3 + y4 + y5}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{y1 ~ b1*x1}
\StringTok{y2 ~ b2*x2 }
\StringTok{y3 ~ b3*x3}
\StringTok{y4 ~ b4*x4}
\StringTok{y5 ~ b5*x5}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ x1 + x2 + x3 + x4 + x5}
\StringTok{# Alternatively: constrain all to 0 for RE model, or}
\StringTok{# just individual correlations}
\StringTok{# a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5}
\StringTok{x1 ~~ x2 + x3 + x4 + x5}
\StringTok{x2 ~~ x3 + x4 + x5}
\StringTok{x3 ~~ x4 + x5}
\StringTok{x4 ~~ x5}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{y1 ~~ e1*y1}
\StringTok{y2 ~~ e2*y2}
\StringTok{y3 ~~ e3*y3}
\StringTok{y4 ~~ e4*y4}
\StringTok{y5 ~~ e5*y5}
\StringTok{'}
\NormalTok{fe_sem_fullyrelaxed.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem_fullyrelaxed, }
                                \DataTypeTok{data =}\NormalTok{ dfw, }
                                \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{anova}\NormalTok{( fe_sem.fit, re_sem.fit)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{anova}\NormalTok{( fe_sem.fit, fe_semb.fit, fe_semc.fit, fe_semd.fit)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem2 <-}\StringTok{ '}
\StringTok{# Define individual effects variable }
\StringTok{a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{# Now the imprecisely measured indicator tilde\{x\}_kt}
\StringTok{# instead of the true variable x_t}
\StringTok{y1 ~ b*x11 }
\StringTok{y2 ~ b*x12 }
\StringTok{y3 ~ b*x13}
\StringTok{y4 ~ b*x14}
\StringTok{y5 ~ b*x15}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ x11 + x12 + x13 + x14 + x15}
\StringTok{x11 ~~ x12 + x13 + x14 + x15}
\StringTok{x12 ~~ x13 + x14 + x15}
\StringTok{x13 ~~ x14 + x15}
\StringTok{x14 ~~ x15}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{y1 ~~ e*y1}
\StringTok{y2 ~~ e*y2}
\StringTok{y3 ~~ e*y3}
\StringTok{y4 ~~ e*y4}
\StringTok{y5 ~~ e*y5}
\StringTok{'}
\NormalTok{fe_sem2.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem2, }
                    \DataTypeTok{data =}\NormalTok{ dfw, }
                    \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{( fe_sem2.fit)}
\end{Highlighting}
\end{Shaded}

@@

Indeed, the magnitude of the bias in a multivariate model is somewhat
more complex to derive, but it will be \begin{align}
\hat{\beta} & = \beta \frac{\mathop{\mathrm{\mathrm{Var}}}(\theta)}{\mathop{\mathrm{\mathrm{Var}}}(\theta) + \mathop{\mathrm{\mathrm{Var}}}(\upsilon)}
\end{align} where \(\theta\) is just the residual of a regression in
which the true underlying variable is regressed on all other covariates.
In this case, we need to regress \(x_{t}\) on \(\alpha_{1}\) and
\(\alpha_{2}\) for:
\(x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}\)
where \(\tau\) is the intercept, and \(\gamma_{1}\), \(\gamma_{2}\) are
the regression coefficients and \(\theta_{t}\) is the residual
\citep[p.~318--320]{Wooldridge2009}.

Normally it is not possible to reconstruct the bias since in cases where
we have to rely on indicators, we would not know the true underlying
variable. Furthermore, in the case of a fixed-effects model, the
covariates are the unobserved time-invariant characteristics. However,
because we are working with simulated data, we have everything we need.
Going back to the results above, we can get the residuals of \(x_{t}\)
by either running a regression and saving the residuals, or we could
skip a step and get them directly using the `residual maker' matrix
\citep{Ruettenauer2020} which is
\(\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\)
and
\(\bm{A} = \begin{pmatrix}\bm{\iota}_{n} & \bm{\alpha_{1}} & \bm{\alpha_{2}}\end{pmatrix}\)
is the \(n \times 3\) matrix of the covariates (plus a constant).

>>>>

\hypertarget{model-notation}{%
\section{Model notation}\label{model-notation}}

The following is an outline of matrix notation for those that prefer it
to the path diagrams and model syntax. There are a number of different
model notations (see, for example \citet{Bollen1989} for an overview),
but the one that will serve us best is one that was proposed by
\citet{Graff1979}: \begin{align}
\bm{y}^{*} & = \bm{\Lambda_{y}}^{*} \bm{\eta}^{*}, \\
\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*}, 
\end{align} where
\(\bm{\eta}^{*} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}\),
\(\bm{\zeta}^{*} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}\),
\(\bm{y}^{*} = (\bm{y}, \bm{x})^{\intercal}\). \(\bm{y}\) is a vector of
observed dependent variables and \(\bm{x}\) is a vector of observed
independent variables. \(\bm{\eta}\) is a vector of the latent dependent
variables and \(\bm{\xi}\) is a vector of latent independent variables.
\(\bm{\varepsilon}\) and \(\bm{\delta}\) are vectors of the errors of
the observed dependent and independent variables, respectively, and
\(\bm{\zeta}\) is a vector of the errors, or disturbances, of the latent
variables. Notice the \(^{*}\) symbol is just meant to differentiate the
vectors with them from those without them. That means, \(\bm{\eta}^{*}\)
is a vector that holds the observed and latent variables, both dependent
(in SEM they are referred to as `endogenous') and independent (i.e.,
`exogeneous')\footnote{For our purposes, the terms endogenous and
  dependent, on the one hand, and exogenous and independent, on the
  other, can be used interchangeably.}, \(\bm{\zeta}^{*}\) holds the
errors for the observed variables and the disturbances of the latent
variables. \(\bm{y}^{*}\) holds just the observed variables, both
dependent and independent, and \(\bm{\Lambda_{y}}^{*}\) is a matrix of
ones and zeros that selects the observed variables from
\(\bm{\eta}^{*}\). Lastly, \(\bm{B}\) is a matrix that holds the
regression coefficients.\footnote{If we say that \(p\) and \(q\) stand
  for the number of observed dependent and independent variables,
  respectively, and \(m\) and \(n\) stand for the number of latent
  dependent and independent variables, respectively, then
  \(\bm{\eta}^{*}\) and \(\bm{\zeta}^{*}\) are \(p + q + m + n\),
  \(\bm{y}^{*}\) is \(p + q\), \(\bm{\Lambda_{y}}^{*}\) is
  \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
  \((p + q + m + n) \times (p + q + m + n)\) \citep{Bollen1989}.}

This notation may be confusing at first, but it has advantages. First,
it allows us the flexibility we need for the models. For example, it
allows observed \(\bm{x}\) to directly influence observed \(\bm{y}\)
(more common notation assumes that substantive effects occur only
between latent variables, observed ones are only used as indicators, see
for example \citet{Bollen1989}, \citet{Kline2016}). It also allows
\(\bm{\xi}\), i.e., any latent exogenous variables, to influence
\(\bm{y}\) directly. These two scenarios cover the traditional FE model
with observed variables, and one in which latent variables are used to
account for measurement error in the independent variables. It is also
consistent with the notation used for these models in \texttt{lavaan}.
In fact, \texttt{lavaan} switches automatically between matrix notations
depending on the specified model. That means the matrix representation
of the model one sees if they type in
\texttt{lavInspect(model,\ what\ =\ "est")} after specifying their model
in \texttt{lavaan} will match up with the notation used here.\footnote{In
  \texttt{Mplus}, the model matrices can be requested by including
  \texttt{OUTPUT:\ TECH1} in the input file.} It does have a potential
disadvantage however. Besides being less intuitive than the typical
\(\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}\)
notation, it means that by including the observed covariates in the
stacked long vector \(\bm{y}^{*}\), they are treated as another response
variable with variances and covariances to be estimated by the model
(instead of just using the sample statistics). This means that the
assumption of multivariate normality (otherwise just imposed on the
dependent variables) also applies to the independent ones
\citep[p.~75]{Skrondal2004}. This can be problematic for noncontinuous
independent variables like sex, nationality dummies, marriage status
(married/unmarried), etc. See \citet{Skrondal2004} for more on this
topic.

Let us, however, make things more concrete and take a look at a simple,
three-wave version of the typical FE-SEM using this notation (shown
graphically in Figure \ref{fig:fesem}). For that, we have the following
matrix notation (with labels on the outside of the matrices):
\begin{align}
\begin{split}
\bm{y}^{*} & = \bm{\Lambda_{y}}^{*}\bm{\eta}^{*} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
 y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
 y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
 x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
 x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
 x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix},
\end{split} \label{eq:y}
\end{align} \begin{align}
\begin{split}
\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} Notice in \(\bm{\zeta}^{*}\), for the independent variables
we could either write, for example \(x_{t}\) or \(\delta_{t}\). As
mentioned above, this is due to the model notation treating the
independent variables like dependent variables with
variances/covariances to be estimated. For the sake of simplicity, we
will ignore this subtlety and refer to the observed variable from now
on, keeping in mind that if the multivariate normality assumption holds,
the estimated statistics will likely be sufficiently close to the sample
ones for it to not make much of a difference.

Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like
much yet. We can remedy this by first putting the equation for
\(\bm{\eta}^{*}\) in reduced form, i.e., by getting rid of the dependent
variable on the r.h.s.: \begin{align}
\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
\bm{\eta}^{*} - \bm{B}\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
(\bm{I} - \bm{B})\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
\bm{\eta}^{*} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*},
\end{align} where \(\bm{I}\) is the identity matrix. By substituting
this back into the equation for the observed variables we get
\(\bm{y}^{*} = \bm{\Lambda_{y}}^{*}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*}]\),
which works out to: \begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align} which is of course exactly what we should expect given
Equation \eqref{eq:fe}\footnote{We can use the \texttt{sympy} package in
  \texttt{python} to verify and show the steps for this and other
  examples, see the
  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py}{supplementary
  materials} for the code.}.

\hypertarget{assumptions}{%
\subsection{Assumptions}\label{assumptions}}

What essentially differentiates an FE from an RE model is our assumption
concerning the relationship between the unobserved individual effects
and the model covariates \citep{Bollen2010}. The FE model assumes that
\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] \ne 0\). As such, if we
fail to control for the correlation of the covariate and the
time-invariant part of the error, then the coefficient of interest, here
\(\beta\), will be biased. Our assumption regarding whether the
individual effects are correlated with the model covariates occurs in
\(\mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] = \bm{\Psi}\),
the covariance matrix of the errors \begin{align}
\bm{y}^{*}\bm{y}^{* \intercal} & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})^{\intercal}] \\
 & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\zeta}^{* \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal})] \\
 & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal} \\
 & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal}.
\end{align} In the case of an FE model, \(\bm{\Psi}\) will reflect our
belief that the individual effects are correlated with the model
covariates, here again for demonstration the three-wave model:
\begin{align}
\bm{\Psi} & = \mathop{\mathrm{\mathbb{E}}}
\begin{blockarray}{cccccccc}
 & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
 \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
 \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
 x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
 x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
 x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
 \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
 \end{block}
\end{blockarray}.
\end{align} Knowing this, we can work out the equation for the
coefficient of interest, \(\beta\). For the sake of simplicity, assume
here and throughout mean-centered variables: \begin{align}
\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) & = \mathop{\mathrm{\mathbb{E}}}[y_{t}x_{t}] \\
 & = \mathop{\mathrm{\mathbb{E}}}[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
 & = \mathop{\mathrm{\mathbb{E}}}[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
 & = \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t}) + \beta \mathop{\mathrm{\mathrm{Var}}}(x_{t}) \\
\hat{\beta}_{FE-SEM} & = \frac{\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) - \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t})}{\mathop{\mathrm{\mathrm{Var}}}(x_{t})}. 
\end{align} This should make intuitive sense. From the observed
covariance between the dependent and the independent variable, we are
partialling out the part that is due to the covariance between the
independent variable and the individual effects per unit, and then
dividing by the variance of the independent variable, as usual. For the
RE model, we assume \(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] = 0\)
and the equation reduces to
\(\hat{\beta}_{RE-SEM} = \mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t})/\mathop{\mathrm{\mathrm{Var}}}(x_{t})\).
The rest of the model-implied covariance matrix results from
\(\bm{y}^{*}\bm{y}^{* \intercal}\).

@@

\CommentTok{\# Run the FE model in plm }
\NormalTok{fe1 \textless{}{-}}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x, }
           \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"within"}\NormalTok{, }
           \DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(fe1)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Generate the cluster means for x per id }
\NormalTok{clusterMeanx \textless{}{-}}\StringTok{ }\KeywordTok{aggregate}\NormalTok{(df}\OperatorTok{$}\NormalTok{x, }\DataTypeTok{by =} \KeywordTok{list}\NormalTok{(df}\OperatorTok{$}\NormalTok{id), }\DataTypeTok{FUN =}\NormalTok{ mean)}
\CommentTok{\# Rename the columns }
\KeywordTok{names}\NormalTok{(clusterMeanx) \textless{}{-}}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"xbar"}\NormalTok{)}

\CommentTok{\# Add the cluster means back into df  }
\NormalTok{df \textless{}{-}}\StringTok{ }\KeywordTok{merge}\NormalTok{(df, clusterMeanx, }\DataTypeTok{by =} \StringTok{"id"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe2 \textless{}{-}}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar, }
           \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"random"}\NormalTok{,}
           \DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(fe2)}
\end{Highlighting}
\end{Shaded}

@@

\CommentTok{\# Run the mixed model in lmer with the cluster means for x }
\NormalTok{mixed1 \textless{}{-}}\StringTok{ }\KeywordTok{lmer}\NormalTok{(y }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{id), }\DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(mixed1)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Print the variance components for the plm model  }
\KeywordTok{print}\NormalTok{(}\KeywordTok{ercomp}\NormalTok{(fe2), }
      \DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Print the variance components for the lmer model }
\KeywordTok{print}\NormalTok{(}\KeywordTok{VarCorr}\NormalTok{(mixed1), }
      \DataTypeTok{comp =} \KeywordTok{c}\NormalTok{(}\StringTok{"Variance"}\NormalTok{, }\StringTok{"Std.Dev"}\NormalTok{), }
      \DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe\_sem\_fullyrelaxed \textless{}{-}}\StringTok{ \textquotesingle{}}
\StringTok{\# Define individual effects variable }
\StringTok{a =\textasciitilde{} y1 + y2 + y3 + y4 + y5}
\StringTok{\# Regressions, constrain coefficient to be equal over time}
\StringTok{y1 \textasciitilde{} b1*x1}
\StringTok{y2 \textasciitilde{} b2*x2 }
\StringTok{y3 \textasciitilde{} b3*x3}
\StringTok{y4 \textasciitilde{} b4*x4}
\StringTok{y5 \textasciitilde{} b5*x5}
\StringTok{\# Allow unrestricted correlation between eta and covariates}
\StringTok{a \textasciitilde{}\textasciitilde{} x1 + x2 + x3 + x4 + x5}
\StringTok{\# Alternatively: constrain all to 0 for RE model, or}
\StringTok{\# just individual correlations}
\StringTok{\# a \textasciitilde{}\textasciitilde{} 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5}
\StringTok{x1 \textasciitilde{}\textasciitilde{} x2 + x3 + x4 + x5}
\StringTok{x2 \textasciitilde{}\textasciitilde{} x3 + x4 + x5}
\StringTok{x3 \textasciitilde{}\textasciitilde{} x4 + x5}
\StringTok{x4 \textasciitilde{}\textasciitilde{} x5}
\StringTok{\# Constrain residual variances to be equal over time}
\StringTok{y1 \textasciitilde{}\textasciitilde{} e1*y1}
\StringTok{y2 \textasciitilde{}\textasciitilde{} e2*y2}
\StringTok{y3 \textasciitilde{}\textasciitilde{} e3*y3}
\StringTok{y4 \textasciitilde{}\textasciitilde{} e4*y4}
\StringTok{y5 \textasciitilde{}\textasciitilde{} e5*y5}
\StringTok{\textquotesingle{}}
\NormalTok{fe\_sem\_fullyrelaxed.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem\_fullyrelaxed, }
                                \DataTypeTok{data =}\NormalTok{ dfw, }
                                \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{anova}\NormalTok{( fe\_sem.fit, re\_sem.fit)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{anova}\NormalTok{( fe\_sem.fit, fe\_semb.fit, fe\_semc.fit, fe\_semd.fit)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe\_sem2 \textless{}{-}}\StringTok{ \textquotesingle{}}
\StringTok{\# Define individual effects variable }
\StringTok{a =\textasciitilde{} 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
\StringTok{\# Regressions, constrain coefficient to be equal over time}
\StringTok{\# Now the imprecisely measured indicator tilde\{x\}\_kt}
\StringTok{\# instead of the true variable x\_t}
\StringTok{y1 \textasciitilde{} b*x11 }
\StringTok{y2 \textasciitilde{} b*x12 }
\StringTok{y3 \textasciitilde{} b*x13}
\StringTok{y4 \textasciitilde{} b*x14}
\StringTok{y5 \textasciitilde{} b*x15}
\StringTok{\# Allow unrestricted correlation between eta and covariates}
\StringTok{a \textasciitilde{}\textasciitilde{} x11 + x12 + x13 + x14 + x15}
\StringTok{x11 \textasciitilde{}\textasciitilde{} x12 + x13 + x14 + x15}
\StringTok{x12 \textasciitilde{}\textasciitilde{} x13 + x14 + x15}
\StringTok{x13 \textasciitilde{}\textasciitilde{} x14 + x15}
\StringTok{x14 \textasciitilde{}\textasciitilde{} x15}
\StringTok{\# Constrain residual variances to be equal over time}
\StringTok{y1 \textasciitilde{}\textasciitilde{} e*y1}
\StringTok{y2 \textasciitilde{}\textasciitilde{} e*y2}
\StringTok{y3 \textasciitilde{}\textasciitilde{} e*y3}
\StringTok{y4 \textasciitilde{}\textasciitilde{} e*y4}
\StringTok{y5 \textasciitilde{}\textasciitilde{} e*y5}
\StringTok{\textquotesingle{}}
\NormalTok{fe\_sem2.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem2, }
                    \DataTypeTok{data =}\NormalTok{ dfw, }
                    \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{( fe\_sem2.fit)}
\end{Highlighting}
\end{Shaded}

@@

Indeed, the magnitude of the bias in a multivariate model is somewhat
more complex to derive, but it will be \begin{align}
\hat{\beta} & = \beta \frac{\mathop{\mathrm{\mathrm{Var}}}(\theta)}{\mathop{\mathrm{\mathrm{Var}}}(\theta) + \mathop{\mathrm{\mathrm{Var}}}(\upsilon)}
\end{align} where \(\theta\) is just the residual of a regression in
which the underlying theoretical variable is regressed on all other
covariates. In this case, we need to regress \(x_{t}\) on \(\alpha_{1}\)
and \(\alpha_{2}\) for:
\(x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}\)
where \(\tau\) is the intercept, and \(\gamma_{1}\), \(\gamma_{2}\) are
the regression coefficients and \(\theta_{t}\) is the residual
\citep[p.~318--320]{Wooldridge2009}.

Normally it is not possible to reconstruct the bias since in cases where
we have to rely on indicators, we would not have observed the underlying
theoretical variable. Furthermore, in the case of a fixed-effects model,
the covariates are the unobserved time-invariant characteristics.
However, because we are working with simulated data, we have everything
we need. Going back to the results above, we can get the residuals of
\(x_{t}\) by either running a regression and saving the residuals, or we
could skip a step and get them directly using the `residual maker'
matrix \citep{Ruettenauer2020} which is
\(\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\)
and
\(\bm{A} = \begin{pmatrix}\bm{\iota}_{n} & \bm{\alpha_{1}} & \bm{\alpha_{2}}\end{pmatrix}\)
is the \(n \times 3\) matrix of the covariates (plus a constant).
